quicR: An R Package for Real-Time Quaking Induced Conversion (RT-QuIC) Assays

### 2025-01-05 man/manuscript/manuscript.qmd
Updated manuscript.

abstract: Real-time quaking induced conversion (RT-QuIC) has quickly become an emerging diagnostic tool for protein misfolding disorders such as Creutzfeldt-Jakob disease and Parkinson's disease. Given that the technology is still relatively new, academic and industry standards have yet to be established. 'quicR' was developed to fill this lack of standardization by providing functions for both data curation and analysis. 
keywords: quicR, R package, RT-QuIC, prion, CJD, Parkinson's
format: pdf
mainfont: Calibri
bibliography: references.bib
bibliographystyle: apa
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(quicR)
library(dplyr)
library(kableExtra)
library(readxl)
library(readr)
library(ggplot2)
library(reshape2)

file <- "../../inst/extdata/input_files/test3.xlsx"
sample_file <- "../../inst/extdata/BMG_formatting/plate_layout.csv"
```

@@

MARS allows input of a TXT file containing sample IDs, dilution factors, and their well locations. This file is uniquely formatted, and not easily reproduced manually. The function "BMG_format" allows for input of a CSV file containing the plate layout, and exports the formatted TXT file.

### Example Plate Layout

\begin{table}[ht]
\centering
\begin{tabular}{|c|cccccccccccc|}
  \hline
   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\ 
  \hline
  A & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  B & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  C & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  D & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  E & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  F & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  G & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  H & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  \hline
\end{tabular}
\end{table}

@@

```{r, message=FALSE}
BMG_format(sample_file, write_file = FALSE)[1:18] %>% 
  as.data.frame()
```

@@

```{r, message=FALSE}
tabs <- organize_tables(file)[-1]
tabs

tabs <- convert_tables(tabs)
tabs %>% head(15) %>% kable(row.names = FALSE)
```

@@

```{r}
df_ <- get_real(file, ordered = FALSE)[[1]] %>% as.data.frame()

df_[1:15, 1:6] %>% kable(row.names = FALSE)
```

@@

```{r}
transpose_real(df_)[1:15, 1:13] %>% kable(row.names = FALSE)
```

@@

```{r}
df_norm <- normalize_RFU(df_, bg_cycle = 4, transposed = FALSE)

df_norm[1:15, 1:12] %>%
  mutate_at(2:ncol(.), ~round(as.numeric(.), 2)) %>%
  kable(row.names = FALSE)
```

@@

## Calculations
There are three analytical metrics with dedicated functions: time-to-threshold, maxpoint ratio, and maximum slope. The rate of amyloid formation does not have a designated function since it is simply the inverse of the time-to-threshold. Each function below accepts input from the "transpose_real" or the "normalize_RFU" functions.

### Time-to-threshold
Time-to-threshold is calculated using the "calculate_TtT" function. The function must be supplied a threshold; default value is 2 (i.e. twice the background fluorescence if the data is normalized). A starting column should also be given as an integer; default value is 3. This is essentially asking how many columns of metadata are included before the fluorescence reads begin.

Time-to-threshold is calculated by iterating through each row and checking if a value is greater than the threshold. If the value is greater, the slope of the previous timepoint to the current timepoint is calculated, and the time intersection of the current read is returned.

### Maxpoint Ratio
Maxpoint ratios (MPR) are calculated by the "calculate_MPR" function. Data must be normalized in order to derive this metric. In a normalized data set, the MPR is simply the maximum value achieved during the run.

### Maximum Slope
The Maximum Slope (MS) is calculated by the "calculate_MS" function. The data can be either raw or normalized. The function iterates through each row using a rolling window which can be adjusted (default value is 3). Given the window size, the slope is calcualated based on the range of the window. The MS is simply the largest slope value recorded.

```{r}
samples <- tabs$`Sample IDs` %>% na.omit()
dilutions <- tabs$Dilutions %>% na.omit() %>% as.numeric()

# Define the number of hours that the rxn ran for.
hours <- as.numeric(colnames(df_norm)[ncol(df_norm)])

df_analyzed <- data.frame("Sample IDs" = samples, check.names = FALSE) %>%
  mutate(
    Dilutions = -log10(dilutions),
    # Maxpoint Ratio
    MPR = calculate_MPR(df_norm, start_col = 3, data_is_norm = TRUE),
    # Max Slope
    MS = calculate_MS(df_norm, data_is_norm = TRUE),
    # Time to Threshold
    TtT = calculate_TtT(df_norm, threshold = 2, start_col = 3),
    # Rate of Amyloid Formation
    RAF = ifelse(TtT == hours, 0, 1 / TtT)
  )

df_analyzed%>%
  head(15) %>%
  kable(row.names = FALSE)
```

@@

### Plate View
```{r, warning=FALSE}
#| fig-width: 8
#| fig-height: 7
sample_locations <- get_sample_locations(
  file, 
  dilution_bool = TRUE,
  dilution_fun = function(x) -log10(x)
)

@@

  ggplot(aes(`Sample IDs`, value, fill = Dilutions)) +
    geom_boxplot() +
    facet_wrap(~variable, scales = "free")
```

@@

>>>>

abstract: Real-time quaking induced conversion (RT-QuIC) has quickly become an emerging diagnostic tool for protein misfolding disorders such as Creutzfeldt-Jakob disease and Parkinson's disease. Given that the technology is still relatively new, academic and industry standards have yet to be established. 'quicR' was developed to fill this lack of standardization by providing functions for both data curation and analysis. 
keywords: quicR, R package, RT-QuIC, prion, CJD, Parkinson's
format: 
  pdf:
    papersize: letter
    documentclass: article
    fontsize: 12pt
    geometry:
      - top=30mm
      - right=30mm
      - bottom=30mm
      - left=30mm
mainfont: Calibri
bibliography: references.bib
bibliographystyle: apa
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(quicR)
library(dplyr)
library(kableExtra)
library(readxl)
library(readr)
library(ggplot2)
library(reshape2)

file <- "../../inst/extdata/input_files/test2.xlsx"
file_384 <- "../../inst/extdata/input_files/test384.xlsx"
sample_file <- "../../inst/extdata/BMG_formatting/plate_layout.csv"
```

@@

MARS allows input of a TXT file containing sample IDs, dilution factors, and their well locations. This file is uniquely formatted, and not easily reproduced manually. The function, "BMG_format", allows for input of a CSV file containing the plate layout, and exports the formatted TXT file.

### Example CSV File Plate Layout

\begin{table}[ht]
\centering
\begin{tabular}{|c|cccccccccccc|}
  \hline
   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\ 
  \hline
  A & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  B & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  C & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  D & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  E & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  F & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  G & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  H & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  \hline
\end{tabular}
\end{table}

@@

```{r, message=FALSE}
BMG_format(sample_file, write_file = FALSE)[1:12] %>%
  as.data.frame()
```

@@

```{r, message=FALSE}
tabs <- organize_tables(file)[-2]
```

#### Sample IDs
```{r, echo=FALSE, as.is=TRUE}
rename <- function(x) {
  names(x) <- 1:12
  rownames(x) <- LETTERS[1:8]
  return(x)
}

tabs$`Sample IDs` %>%
  as.data.frame() %>%
  rename() %>%
  kable(align = rep("c", 12))
```

#### Dilutions
```{r, echo=FALSE, as.is=TRUE}
tabs$Dilutions %>%
  as.data.frame() %>%
  rename() %>%
  kable(align = rep("c", 12))
```
```{r, message=FALSE}
tabs <- convert_tables(tabs)
tabs %>% head(12) %>% kable(row.names = FALSE)
```

#### Sample Locations
Samples locations can be extracted based on their well ID. The "get_sample_locations" function accepts additional arguments if dilution factors were exported from MARS. The "dilution_fun" argument will supply a function for transforming the dilution factors (e.g. if the user would want to perform a log transformation).
```{r, eval=FALSE}
get_sample_locations(
  file, dilution_bool = TRUE, dilution_fun = function(x) -log10(x), sep = " "
)
```
```{r, echo=FALSE}
get_sample_locations(
  file, dilution_bool = TRUE, dilution_fun = function(x) -log10(x), sep = " "
) %>%
  head(12) %>%
  kable(row.names = FALSE, align = c("c", "c"), escape = TRUE)
```

@@

```{r, eval=FALSE}
get_real(file)[[1]]
```
```{r, echo=FALSE}
columns <- 8
df_ <- get_real(file, ordered = FALSE)[[1]] %>% as.data.frame()
df_[1:11, 1:columns] %>% kable(row.names = FALSE, align = rep("c", columns))
```

@@

```{r, eval=FALSE}
get_real(file)[[1]] %>% transpose_real()
```
```{r, echo=FALSE}
columns <- 12
transpose_real(df_)[1:7, 1:columns] %>% 
  kable(row.names = FALSE, align = rep("c", columns))
```

@@

```{r, eval=FALSE}
get_real(file)[[1]] %>% normalize_RFU(transposed = FALSE)
```
```{r, echo=FALSE, message=FALSE}
df_norm <- normalize_RFU(df_, bg_cycle = 4, transposed = FALSE)

df_norm[1:7, 1:columns] %>%
  mutate_at(2:ncol(.), ~round(as.numeric(.), 2)) %>%
  kable(row.names = FALSE, align = rep("c", columns))

df_norm <- df_norm %>%
  mutate(`Sample IDs` = tabs$`Sample IDs`)
```

@@

## Calculations
There are three analytical metrics with dedicated functions: time-to-threshold (TtT), maxpoint ratio (MPR), and maximum slope (MS). The rate of amyloid formation does not have a designated function since it is simply the inverse of the time-to-threshold. Each function below accepts input from the "transpose_real" or the "normalize_RFU" functions.

### Time-to-threshold
TtT is calculated using the "calculate_TtT" function. The function must be supplied a threshold; default value is 2 (i.e. twice the background fluorescence if the data is normalized). A starting column should also be given as an integer; default value is 3. This is essentially asking how many columns of metadata are included before the fluorescence reads begin.

TtT is calculated by iterating through each row and checking if a value is greater than the threshold. If the value is greater, the slope of the previous time-point to the current time-point is calculated, and the time intersection of the current read is returned.

### Maxpoint Ratio
MPR is calculated by the "calculate_MPR" function. Data must be normalized in order to derive this metric. In a normalized data set, the MPR is simply the maximum value achieved during the run. Raw data can be passed to this function, but the argument, "data_is_norm", must be set to TRUE. This will pass the raw data to "normalize_RFU" before calculating the MPR values.

### Maximum Slope
MS is calculated by the "calculate_MS" function. The function iterates through each row using a rolling window which can be adjusted (default value is 3). Given the window size, the slope is calcualated based on the range of the window. The MS is simply the largest slope value recorded.

```{r, eval=FALSE}
samples <- tabs$`Sample IDs`
dilutions <- tabs$Dilutions

df_norm <- get_real(file) %>% normalize_RFU()

data.frame("Sample IDs" = samples, check.names = FALSE) %>%
  mutate(
    Dilutions = -log10(dilutions),
    # Maxpoint Ratio
    MPR = calculate_MPR(df_norm, start_col = 3, data_is_norm = TRUE),
    # Max Slope
    MS = calculate_MS(df_norm, data_is_norm = TRUE),
    # Time to Threshold
    TtT = calculate_TtT(df_norm, threshold = 2, start_col = 3),
    # Rate of Amyloid Formation
    RAF = 1 / TtT
  )
```

```{r, echo=FALSE}
samples <- tabs$`Sample IDs`
dilutions <- tabs$Dilutions %>% na.omit() %>% as.numeric()

df_analyzed <- data.frame("Sample IDs" = samples, check.names = FALSE) %>%
  mutate(
    Dilutions = -log10(dilutions),
    # Maxpoint Ratio
    MPR = calculate_MPR(df_norm, start_col = 3, data_is_norm = TRUE),
    # Max Slope
    MS = calculate_MS(df_norm, data_is_norm = TRUE),
    # Time to Threshold
    TtT = calculate_TtT(df_norm, threshold = 2, start_col = 3),
    # Rate of Amyloid Formation
    RAF = 1 / TtT
  )

df_analyzed%>%
  head(12) %>%
  kable(row.names = FALSE, align = rep("c", ncol(df_analyzed)))
```

@@

### Plate View
The "plate_view" function requires *un-transposed* data and sample locations as arguments. It also has an argument for plate type which will either be 96 or 384.
```{r, warning=FALSE}
#| fig-width: 12
#| fig-height: 8
sample_locations <- get_sample_locations(
  file, 
  dilution_bool = TRUE,
  dilution_fun = function(x) -log10(x)
)

@@

  ggplot(aes(`Sample IDs`, value, fill = Dilutions)) +
    geom_boxplot() +
    facet_wrap(~variable, scales = "free") +
    theme(
      legend.position = "bottom",
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )
```

@@

## Installation
```{r, eval=FALSE}
# Latest CRAN release
install.packages("quicR")

# Development version
install.packages("devtools")
library(devtools)

install_github("gage1145/quicR")
```
